Concurrent Programming in Harmony

### 2021-02-04 book/paper.tex
Moved linked list to concurrent data structures

The most common use for locks is in building concurrent data structures.
By way of example, we will demonstrate how to build a concurrent queue.
The \texttt{queue} module will have the following API:
\begin{itemize}
\item $q = \mathtt{Queue}()$: allocate a new queue;
\item $\mathtt{put}(q, v)$: add $v$ to the tail of the queue;
\item $r = \mathtt{get}(q)$: returns $r = \mathtt{None}$
if $q$ is empty or $r = v$ if $v$ was at the head of the queue.
\end{itemize}
See \autoref{fig:queuetest} for a simple test program that uses the queue.

@@

\section*{Exercises}
\begin{problems}
\item \label{ex:qcontains} Add a method $\mathtt{contains}(q, v)$ to \autoref{fig:queue}
that checks to see if $v$ is in queue $q$.
\item Add a method $\mathtt{length}(q)$ to \autoref{fig:queue} that returns the length of the given queue.
The complexity of the method should be $O(1)$, which is to say that you should maintain
the length of the queue as a field member and update it in \texttt{put} and \texttt{get}.
\item \label{ex:qlength} Write a method $\mathtt{check}(q)$
that checks the integrity of the queue in \autoref{fig:queue}.
In particular, it should check the following integrity properties:
\begin{itemize}
\item If the list is empty, $q$$\rightarrow$$\mathtt{tail}$ should be \texttt{None}.
Otherwise, the last element in the linked list starting from $q$$\rightarrow$$\mathtt{head}$
should equal $q$$\rightarrow$$\mathtt{head}$;
Moreover, $q$$\rightarrow$$\texttt{tail}$$\rightarrow$$next$ should be \texttt{None}.
\item The length field that you added in \autoref{ex:qlength} should equal the length
of the list.
\end{itemize}
Method $\mathtt{check}(q)$ should not obtain a lock; instead add the following line
just before releasing the lock in \texttt{put} and \texttt{get}:
\begin{code}
\textbf{assert} \texttt{check()}
\end{code}
\item Add a method $\mathtt{remove}(q, v)$ to \autoref{fig:queue}
that removes all occurrences of $v$, if any, from queue $q$.
\item The test program in \autoref{fig:queuetest} is not very thorough.  Design
and implement a better one.
\item Create a thread-safe sorted binary tree.  Implement a module
\texttt{bintree} with methods
$\mathtt{BinTree}()$ to create a new binary tree, $\mathtt{insert}(t, v)$ that inserts $v$ into
tree $t$, and $\mathtt{contains}(t, v)$ that checks if $v$ is in tree $t$.  Use a single
lock per binary tree.
\item Bonus problem: create a binary tree that uses, instead of a single lock per tree,
a lock for each node in the tree.
\end{problems}

@@

There are also various algorithms where read-only access is wait-free
but updates do require a lock---these are useful in situations where
most accesses are read-only and the performance of updates is less
critical. %~\cite{}.
\autoref{fig:linkedlist} implements an ordered linked
list of integers without duplicates.
(\autoref{fig:lltest} contains test code.)
Values can be added using \texttt{insert}.
There is also a read-only operation: \texttt{contains} checks if
a particular value is in the list.
\glsadd{linearizable}%

The read-only operation \texttt{contains} is wait\_free: it scans
through the list without having to obtain a lock.
Nonetheless, the implementation of the list is
\emph{linearizable}~\cite{HW90},
\index{linearizable}%
a strong notion of consistency that
makes it appear as if each of the operations executes atomically at
some point between their invocation and return.

The list has two ``book-end'' nodes with values \texttt{-inf} and
\texttt{inf} (similar to the Python \texttt{math.inf} constant).
Each node has a lock, a value, and \textit{next}, a pointer to the next node
(which is \texttt{None} for the final \texttt{inf} node).
The \texttt{insert(\textit{lst}, $v$)} method first
finds and locks two consecutive nodes \textit{before} and \textit{after}
such that $\mathit{before}.\mathtt{data.value} < v \le \mathit{after}.\mathtt{data.value}$.
It does so by performing something called \emph{hand-over-hand locking}.
\index{hand-over-hand locking}%
It first locks the first node, which is the \texttt{-inf} node.
Then, iteratively, it obtains a lock on the next node and release the
lock on the last one, and so on, similar to climbing a tree
hand-over-hand.
The \texttt{contains} method can simply scan through the list
because the \textit{next} pointers are updated atomically in such a
way that they always point to a legal suffix of the list.
An invariant of the algorithm is that at any point in time the
list is ``valid,'' starting with a \texttt{-inf} node and ending
with a \texttt{inf} node.

Determining if an implementation of a concurrent data structure
is linearizable involves finding what are known as the
\emph{linearization points}
\index{linearization point}%
of the operations in an execution.  These are the unique
points in time at which an operation appears to execute atomically.
The linearization point for the \texttt{insert} operation coincides exactly with the update
of the \textit{before}.\texttt{next} pointer.
The linearization point of a \texttt{contains} method
execution depends on whether the value is found or not.
If found, it coincides with retrieving the pointer to the node
that has the value.
If not found, it coincides with retrieving the pointer to the
\texttt{inf} node.

\section*{Exercises}
\begin{problems}
\item Add read-only methods to the data structure in \autoref{fig:linkedlist}
that report the size of the list, the minimum value in the
list, the maximum value in the list, and the sum of the values in the list.
Are they linearizable?  If so, what are their linearization points?
\item A \emph{seqlock}
\index{seqlock}%
consists of a lock and a version number.
An update operation acquires the lock, increments the version number, makes the
changes to the data structure, and then releases the lock.  A read-only operation
does not use the lock.  Instead, it retrieves the version number,
reads the data structure, and then checks if the
version number has changed.  If so, the read-only operation is retried.
Use a seqlock to implement a bank much like \autoref{ex:bank}, with
one seqlock for the entire bank (i.e., no locks on individual accounts).
Method \texttt{transfer} is an update operation; method \texttt{total} is a
read-only operation.  Explain how a seqlock can lead to starvation.
\end{problems}

>>>>

The most common use for locks is in building concurrent data structures.
By way of example, we will first demonstrate how to build a concurrent queue.
The \texttt{queue} module will have the following API:
\begin{itemize}
\item $q = \mathtt{Queue}()$: allocate a new queue;
\item $\mathtt{put}(q, v)$: add $v$ to the tail of the queue;
\item $r = \mathtt{get}(q)$: returns $r = \mathtt{None}$
if $q$ is empty or $r = v$ if $v$ was at the head of the queue.
\end{itemize}
See \autoref{fig:queuetest} for a simple test program that uses the queue.

@@

\index{big lock}%
\index{lock granularity}%
\index{fine-grained lock}%
\index{coarse-grained lock}%
A queue has the nice property that usually only the head or the tail is accessed.
However, in many data structures it is necessary to ``walk'' the data structure,
an operation that can take significant time.
In such a case, a single lock (known as a ``big lock'')
for the entire data structure might restrict concurrency to an unacceptable level.
To reduce the granularity of locking, each node in the data structure must be
endowed with its own lock instead.

\autoref{fig:linkedlist} implements an ordered linked
list of integers without duplicates.
(\autoref{fig:lltest} contains test code.)
Values can be added using \texttt{insert} or deleted using \texttt{remove}.
Method \texttt{contains} checks if a particular value is in the list.

The list has two dummy ``book-end'' nodes with values \texttt{-inf} and
\texttt{inf} (similar to the Python \texttt{math.inf} constant).
An invariant of the algorithm is that at any point in time the
list is ``valid,'' starting with a \texttt{-inf} node and ending
with a \texttt{inf} node.

\index{hand-over-hand locking}%
Each node has a lock, a value, and \textit{next}, a pointer to the next node
(which is \texttt{None} for the final \texttt{inf} node).
The \texttt{\_find(\textit{lst}, $v$)} helper method first
finds and locks two consecutive nodes \textit{before} and \textit{after}
such that $\mathit{before}$$\rightarrow$$\mathtt{data.value} < v \le \mathit{after}$$\rightarrow$$\mathtt{data.value}$.
It does so by performing something called \emph{hand-over-hand locking}.
It first locks the first node, which is the \texttt{-inf} node.
Then, iteratively, it obtains a lock on the next node and release the
lock on the last one, and so on, similar to climbing a tree
hand-over-hand.
Using \texttt{\_find} the \texttt{insert}, \texttt{remove}, and \texttt{contains}
methods are fairly straightforward.

\glsadd{linearizable}%
The implementation of the list is
\emph{linearizable}~\cite{HW90},
\index{linearizable}%
a strong notion of consistency that
makes it appear as if each of the operations executes atomically at
some point between their invocation and return.
Determining if an implementation of a concurrent data structure
is linearizable involves finding what are known as the
\emph{linearization points}
\index{linearization point}%
of the operations in an execution.  These are the unique
points in time at which an operation appears to execute atomically.
The linearization point for the \texttt{insert} operation coincides exactly with the update
of the \textit{before}.\texttt{next} pointer.
The linearization point of a \texttt{contains} method
execution depends on whether the value is found or not.
If found, it coincides with retrieving the pointer to the node
that has the value.
If not found, it coincides with retrieving the pointer to the
\texttt{inf} node.

\section*{Exercises}
\begin{problems}
\item \label{ex:qcontains} Add a method $\mathtt{contains}(q, v)$ to \autoref{fig:queue}
that checks to see if $v$ is in queue $q$.
\item Add a method $\mathtt{length}(q)$ to \autoref{fig:queue} that returns the length of the given queue.
The complexity of the method should be $O(1)$, which is to say that you should maintain
the length of the queue as a field member and update it in \texttt{put} and \texttt{get}.
\item \label{ex:qlength} Write a method $\mathtt{check}(q)$
that checks the integrity of the queue in \autoref{fig:queue}.
In particular, it should check the following integrity properties:
\begin{itemize}
\item If the list is empty, $q$$\rightarrow$$\mathtt{tail}$ should be \texttt{None}.
Otherwise, the last element in the linked list starting from $q$$\rightarrow$$\mathtt{head}$
should equal $q$$\rightarrow$$\mathtt{head}$;
Moreover, $q$$\rightarrow$$\texttt{tail}$$\rightarrow$$next$ should be \texttt{None}.
\item The length field that you added in \autoref{ex:qlength} should equal the length
of the list.
\end{itemize}
Method $\mathtt{check}(q)$ should not obtain a lock; instead add the following line
just before releasing the lock in \texttt{put} and \texttt{get}:
\begin{code}
\textbf{assert} \texttt{check()}
\end{code}
\item Add a method $\mathtt{remove}(q, v)$ to \autoref{fig:queue}
that removes all occurrences of $v$, if any, from queue $q$.
\item The test program in \autoref{fig:queuetest} is not very thorough.  Design
and implement a better one.
\item Create a thread-safe sorted binary tree.  Implement a module
\texttt{bintree} with methods
\item Add methods to the data structure in \autoref{fig:linkedlist}
that report the size of the list, the minimum value in the
list, the maximum value in the list, and the sum of the values in the list.
Are they linearizable?  If so, what are their linearization points?
\item Create a thread-safe sorted binary tree.  Implement a module \texttt{bintree}
with methods
$\mathtt{BinTree}()$ to create a new binary tree, $\mathtt{insert}(t, v)$ that inserts $v$ into
tree $t$, and $\mathtt{contains}(t, v)$ that checks if $v$ is in tree $t$.  Use a single
lock per binary tree.
\item Create a binary tree that uses, instead of a single lock per tree,
a lock for each node in the tree.
\end{problems}

@@

\begin{comment}
There are also various algorithms where read-only access is wait-free
but updates do require a lock---these are useful in situations where
most accesses are read-only and the performance of updates is less
critical. %~\cite{}.
\end{comment}

\section*{Exercises}
\begin{problems}
\item A \emph{seqlock}
\index{seqlock}%
consists of a lock and a version number.
An update operation acquires the lock, increments the version number, makes the
changes to the data structure, and then releases the lock.  A read-only operation
does not use the lock.  Instead, it retrieves the version number,
reads the data structure, and then checks if the
version number has changed.  If so, the read-only operation is retried.
Use a seqlock to implement a bank much like \autoref{ex:bank}, with
one seqlock for the entire bank (i.e., no locks on individual accounts).
Method \texttt{transfer} is an update operation; method \texttt{total} is a
read-only operation.  Explain how a seqlock can lead to starvation.
\end{problems}
